---
title: "Minnesota Timberwolves Analysis"
author: "Tate Goretzki"
date: "2025-04-04"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

##Introduction

- Analysis of the Minnesota Timberwolves during the 2019 season
- Key Findings
- Detailed Analysis of Key Findings
- Analysis of Shots Made By Time Remaining
- Conclusion

##Shot Chart Analysis

- Made Shots = Green, Missed Shots = Blue
- High concentration of shots in the paint and beyond the three-point line
- Low make percentage beyond the three-point line
- Make percentage drastically decreases the further from the basket the shooter is

```{r shot, echo = FALSE}
#Single Year Chart
library(devtools)
devtools::install_github("abresler/nbastatR@aba9179ef644f263387c1536d6ddd26104d79cf4")

library(nbastatR)
library(ggplot2)
library(dplyr)

# I received an error message to increase buffer size when running the first line of shot code
Sys.setenv("VROOM_CONNECTION_SIZE" = 5000000)

shots <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2019)

# NBA Court Dimensions
court_length <- 94
court_width <- 50

shots <- shots %>%
  mutate(
    #Inches to Feet conversion
    locationX = locationX / 10,  
    locationY = locationY / 10,
    
    locationY = ifelse(locationY > (court_length / 2), court_length - locationY, locationY),
    
    shot_made = ifelse(isShotMade, "Made", "Missed")  
  )

#Creating NBA Full Court
draw_full_court <- function() {
  list(
    # Court boundary
    geom_rect(aes(xmin = -court_width/2, xmax = court_width/2, ymin = 0, ymax = court_length), 
              fill = NA, color = "black"),
    
    # Half-court
    geom_segment(aes(x = -court_width/2, y = court_length/2, xend = court_width/2, yend = court_length/2), 
                 color = "black"),
    
    # Center circle
    geom_curve(aes(x = -6, y = court_length/2, xend = 6, yend = court_length/2), curvature = 1, color = "black"),
    
    #Inbound Areas
    geom_rect(aes(xmin = -8, xmax = 8, ymin = 0, ymax = 19), fill = NA, color = "black"),
    geom_rect(aes(xmin = -8, xmax = 8, ymin = court_length - 19, ymax = court_length), fill = NA, color = "black"),
    
    # Free throw
    geom_curve(aes(x = -6, y = 19, xend = 6, yend = 19), curvature = 1, color = "black"),
    geom_curve(aes(x = -6, y = court_length - 19, xend = 6, yend = court_length - 19), curvature = -1, color = "black"),
    
    # Rim and Backboard
    geom_point(aes(x = 0, y = 4), size = 3, color = "black"),  
    geom_point(aes(x = 0, y = court_length - 4), size = 3, color = "black"),  
    geom_segment(aes(x = -3, y = 4.5, xend = 3, yend = 4.5), color = "black", size = 1.5),
    geom_segment(aes(x = -3, y = court_length - 4.5, xend = 3, yend = court_length - 4.5), color = "black", size = 1.5),
    
    # Three-point Lines
    geom_curve(aes(x = -22, y = 0, xend = 22, yend = 0), curvature = -0.9, color = "black"),
    geom_curve(aes(x = -22, y = court_length, xend = 22, yend = court_length), curvature = 0.9, color = "black"),
    
    
    geom_segment(aes(x = -22, y = 0, xend = -22, yend = 14), color = "black"),
    geom_segment(aes(x = 22, y = 0, xend = 22, yend = 14), color = "black"),
    geom_segment(aes(x = -22, y = court_length, xend = -22, yend = court_length - 14), color = "black"),
    geom_segment(aes(x = 22, y = court_length, xend = 22, yend = court_length - 14), color = "black")
  )
}

# Plotting Shots on Court
ggplot(shots, aes(x = locationX, y = locationY, color = shot_made)) +
  draw_full_court() +  
  geom_point(alpha = 0.2, size = 2) +  
  scale_color_manual(values = c("Made" = "green", "Missed" = "blue")) +
  scale_shape_manual(values = c("Made" = 16, "Missed" = 4)) +
  labs(title = "Minnesota Timberwolves Shot Chart", x = "Baseline", y = "Sideline") +
  theme_minimal() +
  coord_fixed(xlim = c(-court_width/2, court_width/2), ylim = c(0, court_length))  


```

##Key Findings

- Higher misses beyond the three-point line, this suggests improving three-point shooting
- Strong scoring in the paint with a high percentage of makes for attempted shots
- Low amount of mid-range shots further reinforcing a three-point and paint scoring strategy
- Defenses may force three point shots with a low make percentage instead of a high scoring percentage in the paint
- Strong performing team, but could improve scoring percentage

##Detailed Analysis of Key Findings --- Shot Distribution

- The Timberwolves had a high distribution of shots on the court with a relatively low scoring percentage
- Improving shooting form, practicing different shots, and consistent practicing can improve the scoring percentage for three-point shots
- The Timberwolves should focus on shooting in the paint with a high scoring percentage
- Focus on keeping shots close to the basket as the closer to the basket the Timberwolves are, the higher scoring percentage

##Detailed Analysis of Key Findings --- Shot Efficiency

- Low scoring percentage beyond the three point line could result from bad spacing, off-ball movement, or lack in three-point scoring development
- High scoring percentage in the paint indicates strong finishing, ball-handling, strong offensive plays, and body positioning
- Work on improving the fundamentals of basketball to improve the scoring percentage for the team

##Detailed Analysis of Key Findings --- Tactical Analysis

- Defenses may be forcing the Timberwolves to take more threes with a low scoring percentage
- Implement offensive strategies such as motion offense, pick-and-rolls, and set plays to create opportunities for shots and drives to the basket
- Improve catch and shoot efficiency for open looks for three point shots
- Utilize offensive strategies that focus on shots in the paint

##Analysis of Shots Made By Time Remaining

- Used the secondsRemaining variable to analyze when shots are taken in the shot clock
- Can help us understand the reasoning behind the low scoring percentage beyond the three point line
- More shots made late in the shot clock are missed rather than made
- All shots close to the half court line are late in the shot clock
- More shots made earlier in the shot clock are closer to the basket
- Three point attempts are made earlier in the shot clock
```{r shot_time, echo = FALSE}
library(devtools)
devtools::install_github("abresler/nbastatR@aba9179ef644f263387c1536d6ddd26104d79cf4")

library(nbastatR)
library(ggplot2)
library(dplyr)

# Increase buffer size to avoid vroom error
Sys.setenv("VROOM_CONNECTION_SIZE" = 5000000)

# Get shot data for Minnesota Timberwolves in the 2019 season
shots <- teams_shots(teams = "Minnesota Timberwolves", seasons = 2019)

# NBA Court Dimensions
court_length <- 94
court_width <- 50

# Prepare the shot data with time categories and shot result
shots <- shots %>%
  mutate(
    # Convert inches to feet
    locationX = locationX / 10,  
    locationY = locationY / 10,
    
    # Flip the Y-axis for a full court view
    locationY = ifelse(locationY > (court_length / 2), court_length - locationY, locationY),
    
    # Determine whether the shot was made or missed
    shot_made = ifelse(isShotMade, "Made", "Missed"),
    
    # Categorize the shots based on seconds remaining in the shot clock
    time_category = case_when(
      secondsRemaining > 15 ~ "Early",    # Early in shot clock (seconds > 15)
      secondsRemaining > 5  ~ "Mid",      # Mid shot clock (seconds between 5 and 15)
      TRUE ~ "Late"                         # Late in shot clock (seconds <= 5)
    )
  )

# Function to draw a full NBA court
draw_full_court <- function() {
  list(
    # Court boundary
    geom_rect(aes(xmin = -court_width/2, xmax = court_width/2, ymin = 0, ymax = court_length), 
              fill = NA, color = "black"),
    
    # Half-court
    geom_segment(aes(x = -court_width/2, y = court_length/2, xend = court_width/2, yend = court_length/2), 
                 color = "black"),
    
    # Center circle
    geom_curve(aes(x = -6, y = court_length/2, xend = 6, yend = court_length/2), curvature = 1, color = "black"),
    
    # Inbound areas
    geom_rect(aes(xmin = -8, xmax = 8, ymin = 0, ymax = 19), fill = NA, color = "black"),
    geom_rect(aes(xmin = -8, xmax = 8, ymin = court_length - 19, ymax = court_length), fill = NA, color = "black"),
    
    # Free throw arcs
    geom_curve(aes(x = -6, y = 19, xend = 6, yend = 19), curvature = 1, color = "black"),
    geom_curve(aes(x = -6, y = court_length - 19, xend = 6, yend = court_length - 19), curvature = -1, color = "black"),
    
    # Rim and backboard
    geom_point(aes(x = 0, y = 4), size = 3, color = "black"),  
    geom_point(aes(x = 0, y = court_length - 4), size = 3, color = "black"),  
    geom_segment(aes(x = -3, y = 4.5, xend = 3, yend = 4.5), color = "black", size = 1.5),
    geom_segment(aes(x = -3, y = court_length - 4.5, xend = 3, yend = court_length - 4.5), color = "black", size = 1.5),
    
    # Three-point lines
    geom_curve(aes(x = -22, y = 0, xend = 22, yend = 0), curvature = -0.9, color = "black"),
    geom_curve(aes(x = -22, y = court_length, xend = 22, yend = court_length), curvature = 0.9, color = "black"),
    geom_segment(aes(x = -22, y = 0, xend = -22, yend = 14), color = "black"),
    geom_segment(aes(x = 22, y = 0, xend = 22, yend = 14), color = "black"),
    geom_segment(aes(x = -22, y = court_length, xend = -22, yend = court_length - 14), color = "black"),
    geom_segment(aes(x = 22, y = court_length, xend = 22, yend = court_length - 14), color = "black")
  )
}

# Plotting the shot chart with time categories
ggplot(shots, aes(x = locationX, y = locationY, color = shot_made, shape = time_category)) +
  draw_full_court() +  
  geom_point(alpha = 0.6, size = 2) +  
  scale_color_manual(values = c("Made" = "green", "Missed" = "red")) +
  scale_shape_manual(values = c("Early" = 16, "Mid" = 17, "Late" = 18)) +  # Different shapes for time categories
  labs(title = "Minnesota Timberwolves Shot Chart (2019)",
       subtitle = "Shots Colored by Made/Missed and Shaped by Time Category",
       x = "Court X (ft)", 
       y = "Court Y (ft)",
       color = "Shot Result", 
       shape = "Time Category") +
  coord_fixed(xlim = c(-court_width/2, court_width/2), ylim = c(0, court_length)) +
  theme_minimal()
```
##Conclusion

- The Timberwolves struggled to score beyond the three point line in 2019
- This could be due to offensive strategies or lack of offensive talent and skill
- The shot clock did affect the scoring percentage of shots beyond the three point line
